Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-05 paper/paper.tex
correctness of liveness analysis

\documentclass[9pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
%\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage [plain,noend,noline,boxed]{algorithm2e}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\usepackage{subfigure}
%\usepackage{program}
%\usepackage{algorithmic}
\usepackage[boxed]{algorithm2e}
\usepackage{makecell}

@@

\begin{lemma}
Consider    a    grammar    $G$    of   the    kind    described    in
Definition~\ref{def:specialgrammar} and a forward access path $\alpha$
consisting  of symbols \acar\  and \acdr\  only. The  decision problem
$\alpha \in \mathscr{L}(G)$ is undecidable.
\end{lemma}

{\em Proof:}  Given a Turing machine  $TM$ and an input  $w$, we shall
construct an unrestricted  grammar $G$ of the kind  described above, so
that $TM$ will halt on $w$ if and only if $\epsilon \in \mathscr{L}(G)$.

We  shall  represent  the  configuration  of  $TM$  at  any  point  as
$w_L(\state,c)w_R$, where $w_L$ is the string on the left of the head,
$w_R$ is the string on the right  of the head, $c$ is the symbol under
the head and \state\ is the current state of the machine. For each
$(\state,c)$ combination of state and symbol, the grammar
will contain a nonterminal \nont{}{c}. Each step of the machine will
correspond to step of derivation based on the grammar.  In addition, we shall maintain
the following invariant between the configuration of $TM$ and the
current sentential form:

\begin{quote}
  If the configuration of the machine is $w_L(\state,c)w_R$, then the
  sentential form will be $\overline{w}_L\nont{}{c}\,w_R$, where
  $\overline{w}_L$ is the same as $w_L$ but with each symbol $d$
  replaced with $\overline{d}$.
\end{quote}

Assume that the TM starts in a state $S_{init}$ with a tape $cw$ and the
head is positioned on the symbol $c$. Then the start non-terminal  of the
grammar $\nont{}{}$ is associated with the production $\nont{}{} \rightarrow \nont{init}{c}w$ and the
sentential form corresponding to the initial configuration is
$\nont{init}{c}w$. Further each transition of the turing machine is modeled by
productions as follows:

\begin{enumerate}
\item A transition $(S_i, c) \rightarrow (S_j,c',L)$ by two productions
  $\nont{i}{c} \rightarrow \acar \nont{j}{\acar}c'$ and $\nont{i}{c}
  \rightarrow \acdr \nont{j}{\acdr}c'$. 
\item A transition $(S_i, c) \rightarrow (S_j,c',R)$ by two productions
  $\nont{i}{c} \rightarrow c \nont{j}{\acar}\bcar$ and $\nont{i}{c}
  \rightarrow c \nont{j}{\acdr}\bcdr$. 
\end{enumerate}
The idea behind these productions is as follows: Assume that the
current sentential form is
\ldots\bcar\bcdr\nont{i}{\acar}\acar\acar\ldots and Starting from $(S_i,
c)$ if the head writes $c'$ and moves to the left, then the symbol
$c'$ is exposed to the right of the tape head. However, we do not know
which symbol that is under the head after the movement. Therefore we
generate two productions assuming that the symbols are \acar\ and
\bcar. If the guess is right, then to remove the symbol to the left in
the sentential form

We circumvent  the problem by over-approximating the  CFG  by
non-deterministic finite state automata (NFA). The NFA are then
simplified using the $\hookrightarrow$ rules. Finally the simplified
NFAs are converted to DFAs.

>>>>

\documentclass[9pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
%\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage [plain,noend,noline,boxed]{algorithm2e}
\usepackage{amssymb}
\usepackage{amsmath}
%\usepackage{amsthm}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\usepackage{subfigure}
%\usepackage{program}
%\usepackage{algorithmic}
\usepackage[boxed]{algorithm2e}
\usepackage{makecell}

@@

\begin{lemma}
Consider    a    grammar    $G$    of   the    kind    described    in
Definition~\ref{def:specialgrammar} and a forward access path $\alpha$
consisting  of symbols \acar\  and \acdr\  only. The  decision problem
$\alpha \in \mathscr{L}(G)$ is undecidable.
\end{lemma} 
\begin{proof}
Given a  Turing machine  and an input  $w\in (1+0)^*$, we  construct a
grammar $G$  such that  the machine will  halt on  $w$ if and  only if
$\epsilon \in  \mathscr{L}(G)$. The grammar includes the  fixed set of
unrestricted productions  in Definition~\ref{def:specialgrammar}.

We  shall   denote  the  configuration   of  the  Turing   machine  as
$w_l(\state,c)w_r$, where $w_l$ is the string to the left of the head,
$w_r$ is the string to the right, $c$ is the symbol under the head and
\state\ is the current state  of the machine.  For each combination of
state  and   symbol  $(\state,c)$,   the  grammar  will   contain  the
non-terminal \nont{}{c}. We shall synchronize each move of the machine
to a  derivation step  using a context  free production,  followed, if
possible, by  a derivation step  using either $\bcar\acar    \rightarrow
\epsilon$ or $\bcdr\acdr    \rightarrow
\epsilon$.
After each synchronization, we shall establish the following invariant
relation between the machine configuration and the  sentential form:

\begin{quote}
  If the configuration of the machine is $w_l(\state,c)w_r$, then the
  sentential form will be $\overline{w}_l\nont{}{c}\,w_r$, where
  $\overline{w}_l$ is the same as $w_l$ but with each symbol $d$ in $w_l$
  replaced by $\overline{d}$.
\end{quote}

Assume that  the TM  starts in a  state $S_\mathit{init}$ with  a tape
$cw$  and the  head   positioned on  the symbol  $c$. Then  the the
sentential  form   corresponding  to  the   initial  configuration  is
$\nont{\mathit{init}}{c}w$ (we can assume  that there is a production
$\nont{}{}  \rightarrow \nont{\mathit{init}}{c}w$, where  \nont{}{} is
the start symbol of  the grammar). Further correspondences between the
Turing machine moves and the grammar productions are as follows:

\begin{enumerate}
\item For each transition  $(S_i, c) \rightarrow (S_j,c',L)$, there are
  two  productions $\nont{i}{c}  \rightarrow  \acar \nont{j}{\acar}c'$
  and $\nont{i}{c} \rightarrow \acdr \nont{j}{\acdr}c'$.
\item For each  transition $(S_i, c) \rightarrow (S_j,c',R)$, there
  are two productions
  $\nont{i}{c} \rightarrow c \nont{j}{\acar}\bcar$ and $\nont{i}{c}
  \rightarrow c \nont{j}{\acdr}\bcdr$. 
\end{enumerate}
The idea behind  the productions is explained with  an example: Assume
that        the        current        sentential        form        is
\bcar\bcdr\nont{i}{\acar}\acar\acar.  Assume that  the  machine has  a
transition  $(S_i,\acar) \rightarrow (S_j,\acdr,  L)$. Since  the next
corresponding step in the derivation  has to be done without any prior
knowledge of whether the symbol to the left of the tape is a \acar\ or
a  \acdr, two  productions are  provided,  and the  invariant will  be
maintained  only   if  the  production   $\nont{i}{\acar}  \rightarrow
\acdr\nont{j}{\acdr}\acdr$  is  chosen  for   the  next  step  in  the
derivation.         This         gives        the        configuration
\bcar\bcdr\acdr\nont{i}{\acdr}\acdr\acar\acar.    Simplification  with
the    production    $\bcdr\acdr    \rightarrow    \epsilon$    yields
\bcar\nont{i}{\acdr}\acdr\acar\acar, which  exactly corresponds to the
changed configuration  of the machine.  Notice carefully that  a wrong
choice breaks the invariant and it cannot be recovered subsequently by
any choice of productions.

After  the Turing machine  has halted,  there are  further ``cleanup''
derivations  that derive  $\epsilon$ only  if the  invariant  has been
maintained  so far.  For every  symbol  $c$, we introduce  a  non-terminal
$\nont{\mathit{final}}{c}$ where $\nont{\mathit{final}}{}$ is the
final state of the Turing machine. We add
productions $\nont{\mathit{final}}{c} \rightarrow \acar
\nont{\mathit{final}}{c}$ and $\nont{\mathit{final}}{c} \rightarrow \acdr
\nont{\mathit{final}}{c}$ for cleaning up the \bcar\ and \bcdr\ symbols on
the left of the head and $\nont{\mathit{final}}{c} \rightarrow 
\nont{\mathit{final}}{c}\bcar$ and $\nont{\mathit{final}}{c} \rightarrow \acdr
\nont{\mathit{final}}{c}\bcdr$ for cleaning up \acar\ and \acdr\ on the
right of the tape head. This completes the reduction.  
\end{proof}

We circumvent the problem  of undecidability by over- approximating the
CFG by non-deterministic finite state automata (NFA). The NFA are then
simplified using  the $\hookrightarrow$ rules.  Finally the simplified
NFAs are converted to DFAs.
